%==============================================================================
\section{Categorical Selection and Accessibility Pathways}
\label{sec:selection}
%==============================================================================

\subsection{The Selection Problem}

In Maxwell's thought experiment, the demon ``selects'' fast molecules to pass through the door. We now analyse what selection means in categorical terms.

\begin{definition}[Categorical Selection]
\label{def:categorical_selection}
A \textbf{categorical selection} is the completion of a specific categorical state $C^* \in [C]_{\text{spatial}}$ from an equivalence class of spatially indistinguishable states.
\end{definition}

\begin{proposition}[Selection as Equivalence Class Reduction]
\label{prop:selection_reduction}
Categorical selection reduces the equivalence class to a singleton:
\begin{equation}
[C]_{\text{spatial}} \xrightarrow{\text{selection}} \{C^*\}
\end{equation}
This is an information-gain process: $\log |[C]_{\text{spatial}}|$ bits of categorical information are specified.
\end{proposition}

\begin{proof}
Before selection, any state in $[C]_{\text{spatial}}$ is possible. After selection, exactly one state $C^*$ is completed. The information gained is:
\begin{equation}
I_{\text{selection}} = \log_2 |[C]_{\text{spatial}}| - \log_2 1 = \log_2 |[C]_{\text{spatial}}|
\end{equation}
For typical gas systems with $|[C]_{\text{spatial}}| \sim 10^6$, this is $\sim 20$ bits per selection. \qed
\end{proof}

\subsection{Accessibility Through Phase-Lock Networks}

\begin{theorem}[Phase-Lock Accessibility]
\label{thm:phase_lock_accessibility}
When categorical state $C_i$ is completed, the accessible states for subsequent completion are determined by phase-lock adjacency:
\begin{equation}
\accessible(C_i) = \{C_j \in \catspace : \exists (m_k, m_l) \in E(\phaselockgraph) \text{ connecting } C_i \text{ to } C_j\}
\label{eq:accessible_states}
\end{equation}
\end{theorem}

\begin{proof}
Categorical transitions require physical mechanism. The mechanisms available are:
\begin{enumerate}
    \item Molecular collisions: transfer energy and phase information
    \item Phase-lock coupling: synchronise oscillatory states
    \item Electromagnetic interaction: modify electronic configurations
\end{enumerate}

All these mechanisms operate through intermolecular interactions. Molecules not connected in $\phaselockgraph$ have negligible interaction strength (below threshold~\eqref{eq:phase_lock_threshold}), hence cannot mediate transitions.

Therefore, accessible states are precisely those reachable through phase-lock network edges. \qed
\end{proof}

\begin{corollary}[Pathway Opening]
\label{cor:pathway_opening}
Completing categorical state $C_i$ ``opens'' pathways to all states in:
\begin{equation}
\text{Pathways}(C_i) = \{C_j : d_{\catspace}(C_i, C_j) < \infty\}
\end{equation}
the connected component of $\catspace$ containing $C_i$.
\end{corollary}

\subsection{The Cascade Effect}

\begin{theorem}[Categorical Cascade]
\label{thm:categorical_cascade}
Selection of a single categorical state $C_1$ initiates a cascade of accessible completions:
\begin{align}
C_1 &\to \accessible(C_1) = \{C_2^{(1)}, C_2^{(2)}, \ldots\} \\
C_2^{(k)} &\to \accessible(C_2^{(k)}) = \{C_3^{(k,1)}, C_3^{(k,2)}, \ldots\} \\
&\vdots
\end{align}
The cascade propagates through the phase-lock network.
\end{theorem}

\begin{proof}
Each completed state makes adjacent states accessible (Theorem~\ref{thm:phase_lock_accessibility}). Completing any accessible state makes its neighbours accessible. This propagation continues until:
\begin{enumerate}
    \item The entire connected component is exhausted, or
    \item Energy/entropy constraints halt the cascade
\end{enumerate}

The structure of the cascade is determined by $\phaselockgraph$ topology. \qed
\end{proof}

\begin{definition}[Cascade Wavefront]
\label{def:cascade_wavefront}
The \textbf{cascade wavefront} at step $n$ is:
\begin{equation}
W_n = \{C \in \catspace : d_{\catspace}(C_1, C) = n\}
\end{equation}
the set of states at categorical distance $n$ from the initial selection.
\end{definition}

\begin{proposition}[Wavefront Propagation]
\label{prop:wavefront_propagation}
The wavefront size evolves according to:
\begin{equation}
|W_{n+1}| = \sum_{C \in W_n} |\accessible(C) \setminus \gamma(t_n)|
\end{equation}
where $\gamma(t_n)$ is the set of already-completed states.
\end{proposition}

\subsection{Selection Without Information}

We now prove the central result: categorical selection requires no external information.

\begin{theorem}[Information-Free Selection]
\label{thm:information_free}
Categorical selection from equivalence class $[C]_{\text{spatial}}$ is determined by phase-lock network topology without external information input. Specifically:
\begin{equation}
C^* = \argmin_{C \in [C]_{\text{spatial}}} d_{\catspace}(C, C_{\text{prev}})
\end{equation}
where $C_{\text{prev}}$ is the previously completed state.
\end{theorem}

\begin{proof}
Consider a system at categorical state $C_{\text{prev}}$ transitioning to spatial configuration $\mathbf{q}_{\text{new}}$.

The accessible categorical states at $\mathbf{q}_{\text{new}}$ are:
\begin{equation}
\accessible(C_{\text{prev}}) \cap [C]_{\text{spatial}}(\mathbf{q}_{\text{new}})
\end{equation}
the intersection of phase-lock accessible states and spatially compatible states.

\textbf{Case 1: Unique accessible state.}
If $|\accessible(C_{\text{prev}}) \cap [C]_{\text{spatial}}| = 1$, selection is deterministic---only one categorical state is reachable.

\textbf{Case 2: Multiple accessible states.}
If multiple states are accessible, physical dynamics (minimising action, maximising entropy production rate) select among them. This selection is governed by:
\begin{equation}
C^* = \argmax_{C \in \accessible(C_{\text{prev}}) \cap [C]_{\text{spatial}}}\frac{d\alpha}{dt}
\end{equation}
where $\alpha$ is the oscillatory termination probability, favouring states with shorter paths to equilibrium.

In both cases, selection follows from network topology and physical dynamics---no external ``measurement'' or ``decision'' is required.

The ``information'' specifying which categorical state to occupy is structural (encoded in $\phaselockgraph$) rather than acquired through measurement. \qed
\end{proof}

\begin{corollary}[No Demon Required]
\label{cor:no_demon}
The selection process attributed to Maxwell's Demon is categorical completion through phase-lock topology. No agent is required because:
\begin{enumerate}
    \item Selection is determined by network structure (Theorem~\ref{thm:information_free})
    \item Accessibility follows from phase-lock adjacency (Theorem~\ref{thm:phase_lock_accessibility})
    \item Cascade propagation is automatic (Theorem~\ref{thm:categorical_cascade})
\end{enumerate}
\end{corollary}

\subsection{Apparent Sorting Through Categorical Pathways}

\begin{theorem}[Apparent Temperature Sorting]
\label{thm:apparent_sorting}
Molecules following categorical pathways appear sorted by temperature because phase-lock clusters correlate with kinetic properties. Specifically, for molecules $i, j$ in the same phase-lock cluster:
\begin{equation}
\text{Cov}(E_{\text{kin},i}, E_{\text{kin},j}) > 0
\label{eq:kinetic_correlation}
\end{equation}
despite phase-lock formation being kinetically independent.
\end{theorem}

\begin{proof}
Phase-lock clusters form based on molecular properties: polarisability, dipole moment, vibrational frequencies. These properties correlate with molecular mass and structure, which in turn correlate with kinetic energy distribution at thermal equilibrium.

Consider the Maxwell-Boltzmann distribution:
\begin{equation}
f(v) = 4\pi \left(\frac{m}{2\pi k_B T}\right)^{3/2} v^2 \exp\left(-\frac{mv^2}{2k_B T}\right)
\end{equation}

Molecules with similar mass $m$ have similar most-probable velocities $v_p = \sqrt{2k_B T/m}$. Since phase-lock clusters tend to contain molecules with similar properties (similar polarisabilities, similar dipole moments), they tend to contain molecules with similar masses, hence similar kinetic energies.

The correlation~\eqref{eq:kinetic_correlation} arises from shared molecular properties, not from kinetic energy determining phase-lock formation.

\textbf{Causal structure:}
\begin{equation}
\text{Molecular properties} \to \begin{cases} \text{Phase-lock clustering} \\ \text{Kinetic energy distribution} \end{cases}
\end{equation}

Both phase-lock structure and kinetic properties are downstream of molecular properties. The correlation is non-causal: neither determines the other. \qed
\end{proof}

\begin{corollary}[Sorting Is Correlation, Not Causation]
\label{cor:correlation_not_causation}
When molecules ``sorted'' by categorical pathways appear to separate by temperature, this reflects:
\begin{enumerate}
    \item Pre-existing phase-lock cluster structure
    \item Correlation between cluster membership and kinetic properties
    \item NOT measurement of velocity followed by sorting decision
\end{enumerate}
\end{corollary}

