
\subsection{Processor Architecture}

The semantic processor integrates molecular encoding, harmonic networks, and atmospheric memory into a complete language understanding system.

\begin{definition}[Semantic Processor]
The semantic processor $\mathcal{P}_{\text{sem}}$ consists of:
\begin{enumerate}
    \item Molecular encoder: Maps words to virtual molecules
    \item Coincidence network: Discovers harmonic relationships
    \item Atmospheric memory: Stores words by S-coordinate
    \item Prediction engine: Computes meaning via structure prediction
\end{enumerate}
\end{definition}

\subsection{Atmospheric Memory}

The atmospheric memory stores word-molecules in a virtual gas, addressable by S-entropy coordinates.

\begin{definition}[Atmospheric Semantic Memory]
The atmospheric memory $\mathcal{A}$ is characterized by:
\begin{itemize}
    \item Virtual volume $V$ (in cm$^3$)
    \item Molecular density $\rho = 2.5 \times 10^{19}$ molecules/cm$^3$
    \item Total molecules $N = \rho \cdot V$
    \item S-resolution $\Delta S$ (categorical precision)
    \item Address count $(1/\Delta S)^3$
\end{itemize}
\end{definition}

\begin{proposition}[Storage Capacity]
For $V = 10$ cm$^3$ and $\Delta S = 0.01$:
\begin{align}
N_{\text{molecules}} &= 2.5 \times 10^{20} \\
N_{\text{addresses}} &= 10^6 \\
N_{\text{molecules/address}} &= 2.5 \times 10^{14}
\end{align}
\end{proposition}

Words are stored at S-coordinates determined by their molecular properties. Retrieval is by categorical address (S-coordinate neighborhood), not by position.

\subsection{Storage Operation}

Storing a word in atmospheric memory:

\begin{algorithmic}[1]
\Function{Store}{word $w$}
    \State Encode molecule: $\mathcal{M} \gets$ encode($w$)
    \State Add to network: network.add($\mathcal{M}$)
    \State Compute address: $\mathbf{a} \gets$ s\_to\_address($\Scoord(\mathcal{M})$)
    \State Store at address: memory[$\mathbf{a}$].append($\mathcal{M}$)
    \State \Return $\mathbf{a}$
\EndFunction
\end{algorithmic}

The word is simultaneously added to the coincidence network (for harmonic discovery) and to the atmospheric memory (for retrieval).

\subsection{Retrieval Operation}

Retrieving words by S-coordinate:

\begin{algorithmic}[1]
\Function{Retrieve}{$\Scoord^*$, radius $r$}
    \State results $\gets \emptyset$
    \State base $\gets$ s\_to\_address($\Scoord^*$)
    \For{each address $\mathbf{a}$ within radius $r$ of base}
        \State results $\gets$ results $\cup$ memory[$\mathbf{a}$]
    \EndFor
    \State \Return results
\EndFunction
\end{algorithmic}

Retrieval returns all word-molecules stored at addresses within the specified radius of the target S-coordinate. This categorical retrieval naturally groups semantically related words.

\subsection{Meaning Prediction}

Predicting the meaning of a word from context:

\begin{algorithmic}[1]
\Function{PredictMeaning}{word $w$, context $C$}
    \State Store context: \textbf{for} $c \in C$ \textbf{do} Store($c$)
    \State Predict frequency: $\omega^*, \text{conf} \gets$ network.predict($w$, $C$)
    \State Compute S-coordinate from context molecules:
    \State \quad $\Sk \gets$ mean($\{\Sk(\mathcal{M}(c)) : c \in C\}$)
    \State \quad $\St \gets$ mean($\{\St(\mathcal{M}(c)) : c \in C\}$)
    \State \quad $\Se \gets$ mean($\{\Se(\mathcal{M}(c)) : c \in C\}$)
    \State Find related: related $\gets$ network.neighbors($C$)
    \State \Return \{word, $\omega^*$, conf, $\Scoord$, related\}
\EndFunction
\end{algorithmic}

The prediction combines frequency prediction from harmonic coincidences with S-coordinate estimation from context.

\subsection{Semantic Understanding}

Complete semantic understanding of text:

\begin{algorithmic}[1]
\Function{Understand}{text $T$, query $Q$}
    \State words $\gets$ tokenize($T$)
    \State addresses $\gets$ [Store($w$) for $w$ in words]
    \For{$w$ in words}
        \State network.add(encode($w$))
    \EndFor
    \If{$Q$ is not null}
        \State predictions $\gets$ [PredictMeaning($q$, words) for $q$ in tokenize($Q$)]
    \Else
        \State predictions $\gets \emptyset$
    \EndIf
    \State completions $\gets$ predict\_next(words)
    \State \Return \{words, predictions, completions, network.stats\}
\EndFunction
\end{algorithmic}

Understanding processes text by:
\begin{enumerate}
    \item Tokenizing into words
    \item Storing each word in atmospheric memory
    \item Building the coincidence network
    \item If query provided, predicting meanings for query words in context
    \item Predicting likely next words (completion)
\end{enumerate}

\subsection{Text Comparison}

Comparing texts using molecular similarity:

\begin{algorithmic}[1]
\Function{Compare}{text$_1$, text$_2$}
    \State $M_1 \gets$ [encode($w$) for $w$ in tokenize(text$_1$)]
    \State $M_2 \gets$ [encode($w$) for $w$ in tokenize(text$_2$)]
    \State $\Scoord_1 \gets$ mean([$\Scoord(\mathcal{M})$ for $\mathcal{M}$ in $M_1$])
    \State $\Scoord_2 \gets$ mean([$\Scoord(\mathcal{M})$ for $\mathcal{M}$ in $M_2$])
    \State $d_S \gets \|\Scoord_1 - \Scoord_2\|$
    \State $\omega_1 \gets$ [$\omega_0(\mathcal{M})$ for $\mathcal{M}$ in $M_1$]
    \State $\omega_2 \gets$ [$\omega_0(\mathcal{M})$ for $\mathcal{M}$ in $M_2$]
    \State overlap $\gets |\text{round}(\log \omega_1) \cap \text{round}(\log \omega_2)|$
    \State similarity $\gets 0.6 \cdot (1/(1+d_S)) + 0.4 \cdot (\text{overlap}/\max(|M_1|, |M_2|))$
    \State \Return similarity
\EndFunction
\end{algorithmic}

Comparison combines S-entropy distance (categorical similarity) with frequency overlap (spectroscopic similarity).

\subsection{Complexity Analysis}

\begin{proposition}[Processing Complexity]
The semantic processor has the following complexities:
\begin{center}
\begin{tabular}{lc}
\toprule
Operation & Complexity \\
\midrule
Word encoding & $O(|w|)$ \\
Network insertion & $O(N \cdot K^2)$ for $N$ existing words, $K$ harmonics checked \\
Storage & $O(1)$ \\
Retrieval (by S-coordinate) & $O(r^3)$ for radius $r$ \\
Meaning prediction & $O(|C| \cdot K^2)$ for context size $|C|$ \\
Full understanding & $O(|T| \cdot N \cdot K^2)$ for text length $|T|$ \\
\bottomrule
\end{tabular}
\end{center}
\end{proposition}

The quadratic factor in network operations is due to checking harmonic coincidences. For fixed $K$ (maximum harmonic order), this is effectively linear in the number of words.

\subsection{Comparison with Attention}

The harmonic coincidence mechanism provides context-dependent processing analogous to attention.

\begin{center}
\begin{tabular}{lcc}
\toprule
Property & Attention & Harmonic Coincidence \\
\midrule
Learned & Yes (QKV weights) & No (intrinsic) \\
Complexity & $O(n^2)$ sequence length & $O(n K^2)$ harmonics \\
Context range & All tokens equally & Harmonic neighbors \\
Computation & Matrix multiplication & Frequency comparison \\
Parameters & $O(d^2)$ per layer & 0 \\
\bottomrule
\end{tabular}
\end{center}

Attention computes weighted combinations with learned weights. Harmonic coincidence identifies related words through frequency relationships, with no learned parameters.

