\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{physics}
\usepackage{graphicx}
\usepackage{float}
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{siunitx}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{natbib}
\usepackage{geometry}
\usepackage{fancyhdr}
\usepackage{tikz}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}

\geometry{margin=1in}
\setlength{\headheight}{14.5pt}
\pagestyle{fancy}
\fancyhf{}
\rhead{\thepage}
\lhead{Resolution of Orgel's Paradox}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}{Example}

\title{\textbf{Thermodynamic Consequences of Geometric Categorical Partitioning in Oscillating Systems: Resolution of Orgel's Paradox through Charge Partitioning Mechanics Trajectories}}

\author{
Kundai Farai Sachikonye\\
\texttt{kundai.sachikonye@wzw.tum.de}
}

\date{\today}

\begin{document}

\maketitle

\begin{abstract}

We present a resolution of Orgel's paradox, which describes the circular dependency between genetic information, enzymes, and metabolism in origin of life scenarios, through the framework of electron transport partitioning. Our analysis demonstrates that the fundamental operation underlying life is not information storage but rather charge separation through electron transport, which constitutes geometric partitioning at the quantum level.

Mathematical analysis establishes that the probability of membrane-first scenarios ($P \approx 10^{-6}$) exceeds that of information-first scenarios (RNA world: $P \approx 10^{-150}$; DNA-first: $P \approx 10^{-200}$) by factors approaching $10^{144}$, representing a difference of over 140 orders of magnitude in thermodynamic feasibility. We prove that autocatalytic electron transport—defined as systems where electron movement creates the conditions necessary for further electron movement—represents the minimal self-referential structure capable of initiating biological complexity without requiring pre-existing information templates.

The universal homochirality of biological molecules, including L-amino acids, D-sugars, and right-handed DNA helices, is shown to be direct evidence of partitioning primacy rather than random symmetry breaking. Electron transport in electromagnetic fields creates chiral selection that propagates through geometric apertures at all organizational levels, from molecular recognition to cellular structure. We establish that biological membranes evolved primarily as electron transport scaffolding rather than as compartmentalization structures, fundamentally reinterpreting their role in the origin of life.

Furthermore, we demonstrate that DNA and RNA evolved as charge capacitors that optimize electrostatic integration within cells, with information storage emerging as an "evolutionary bonus" of charge dynamics rather than as their primary selective function. This framework explains the formation of complex organic molecules in interstellar environments through the semiconductor behavior of mineral surfaces, resolving the long-standing paradox of prebiotic chemistry in cold, irradiated space environments where thermal activation is insufficient.

These results establish electron transport partitioning as the thermodynamically inevitable origin of life, replacing stochastic information-first models with a deterministic framework based on physical necessity. The transition from non-living to living systems emerges as a continuous physical process driven by charge dynamics rather than as a discontinuous event requiring improbable information assembly.



\textbf{Keywords:} origin of life, Orgel's paradox, electron transport, charge partitioning, homochirality, membrane evolution, prebiotic chemistry
\end{abstract}

\section{Introduction}
\label{sec:introduction}


The origin of life remains one of the most challenging and enduring problems in natural science, representing a fundamental gap in our understanding of how non-living matter transitions to living systems. Central to this challenge is Orgel's paradox, which describes the apparent circular dependency between genetic information storage, enzymatic catalysis, and metabolic energy production \citep{orgel1968evolution, orgel2004prebiotic}. This circularity presents a logical impasse: genetic information requires enzymes for replication and expression, enzymes require genetic information for their synthesis, and both require metabolic energy, which itself depends on enzyme-catalyzed reactions. Each component appears to presuppose the existence of the others, creating a chicken-and-egg problem that has resisted resolution for over half a century.

Traditional approaches to resolving this circularity have focused on proposing various ``world'' hypotheses, each attempting to identify a single primordial component from which the others could subsequently emerge. The RNA world hypothesis posits that self-replicating RNA molecules preceded both DNA and proteins, serving simultaneously as genetic material and catalytic agents \citep{gilbert1986origin}. Metabolism-first theories propose that autocatalytic chemical networks arose before genetic polymers, with information storage emerging later to stabilize successful reaction pathways \citep{wachtershauser1988before}. Lipid world scenarios suggest that self-assembling amphiphilic boundaries provided the first compartmentalisation, within which subsequent chemical evolution occurred \citep{segre2001lipid}. However, each of these approaches privileges one component as primordial while leaving fundamentally unexplained how the other essential components arose from this starting point. The RNA world requires an explanation for the spontaneous emergence of catalytically active sequences with sufficient complexity to replicate themselves; metabolism-first scenarios must account for the transition from simple autocatalytic cycles to information-bearing polymers, and lipid world models face the challenge of explaining how compartmentalisation alone leads to the emergence of genetic information and catalysis.

We propose a fundamentally different resolution to Orgel's paradox, one that does not privilege any of the traditional biochemical components. Instead, we argue that the primordial operation underlying life is neither information storage nor metabolism nor compartmentalisation, but rather \emph{electron transport partitioning---the} separation of electrical charge across a geometric boundary through directed electron movement. This operation represents a more fundamental level of organisation than any of the traditionally proposed primordial systems, operating at the interface between quantum mechanics and classical thermodynamics.

The primacy of electron transport partitioning rests on four key principles that distinguish it from information-first or metabolism-first scenarios. First, electron transport inherently creates \emph{partitioning} through charge separation, establishing distinct electrochemical domains on either side of a boundary. This partitioning is the prerequisite for all subsequent categorical operations in living systems, including molecular recognition, selective transport, and the establishment of inside versus outside. Without charge separation, there can be no thermodynamic driving force for the directional processes that characterise life. Second, partitioning through electron transport is \emph{temperature-independent} at the quantum level, operating through tunneling and field effects rather than thermal activation. This property explains the paradoxical observation of complex prebiotic chemistry in cold interstellar environments, where thermal reaction rates would be negligibly slow. Third, autocatalytic electron transport exhibits \emph{self-referential} dynamics, where electron movement creates the electrochemical conditions necessary for further electron movement. This self-referential property allows electron transport systems to perpetuate themselves without requiring external information templates or pre-existing catalytic machinery. Fourth, the universal \emph{homochirality} of biological molecules—the exclusive use of L-amino acids, D-sugars, and right-handed DNA helices across all domains of life—provides direct empirical evidence of partitioning primacy. Chiral selection emerges naturally from electron transport in electromagnetic fields, which create asymmetric potential landscapes that favour one enantiomer over the other through geometric partitioning effects.

This framework fundamentally reinterprets the role of the major biochemical components in the origin and operation of life. Biological membranes, rather than serving primarily as compartmentalisation structures that separate the interior from the exterior, evolved as electron transport scaffolding that optimises charge separation and electrochemical gradient formation. The lipid bilayer provides an insulating barrier with embedded protein complexes that facilitate directed electron flow, creating the charge-separated state that powers all cellular work. Similarly, DNA and RNA, rather than existing primarily as information storage molecules, evolved as charge capacitors that optimise electrostatic integration within the cellular electromagnetic field. The regular helical structure of nucleic acids, their polyanionic backbone, and their association with cationic proteins create charge distributions that stabilise the cellular electric field. Information storage, in this view, emerges as an "evolutionary bonus" of charge dynamics---a secondary function that became possible once charge-stabilising polymers existed, rather than the primary selective pressure driving their evolution.

The electron transport partitioning framework also provides a novel explanation for interstellar prebiotic chemistry, resolving long-standing puzzles about the formation of complex organic molecules in space. Observations of amino acids, nucleobases, and other biomolecular precursors in meteorites, comets, and interstellar dust have demonstrated that prebiotic chemistry occurs in cold, irradiated environments where thermal activation energies are insufficient to drive conventional chemical reactions. We demonstrate that the semiconductor behaviour of mineral surfaces, particularly iron-sulphur minerals and phyllosilicate clays, provides the electron transport scaffolding necessary for charge-driven molecular synthesis. Photoexcitation and cosmic ray bombardment create charge carriers in these semiconducting minerals, and the resulting electric fields drive molecular assembly through geometric partitioning rather than thermal activation.

These results establish electron transport partitioning as the thermodynamically inevitable origin of life, replacing stochastic information-first models with a deterministic framework based on physical necessity. The transition from non-living to living systems emerges not as a discontinuous event requiring the improbable spontaneous assembly of information-bearing polymers, but as a continuous physical process driven by charge dynamics. Life, in this view, is the inevitable consequence of electron transport in structured environments, with information, metabolism, and compartmentalisation emerging as elaborations of the fundamental charge-partitioning operation.


\input{sections/categorical-oscillation-foundation}
\input{sections/orgels-paradox}
\input{sections/electron-transport-charge-partitioning}
\input{sections/autocatalytic-electron-transport}
\input{sections/geometric-partitioning}
\input{sections/homochirality}
\input{sections/electron-transport-scaffolding}
\input{sections/charge-capacitor-evolution}
\input{sections/semiconductor-origins}
\input{sections/electron-transport-to-genome}
\input{sections/genome-structural-element}
\input{sections/spare-tire-principle}

\section{Discussion}
\label{sec:discussion}

The resolution of Orgel's paradox through the framework of electron transport partitioning provides a unified theoretical foundation for understanding the origin of life that differs fundamentally from traditional approaches in both its conceptual structure and its thermodynamic predictions. Rather than treating the emergence of life as a problem of improbable molecular assembly requiring specific sequences of information-bearing polymers, this framework identifies charge separation through electron transport as the primordial operation from which information storage, catalysis, and metabolism subsequently emerged. In this section, we discuss the principal implications of this perspective and examine how it relates to and reinterprets existing theories of life's origin.

\subsection{Thermodynamic Inevitability vs. Stochastic Accident}

Traditional origin of life theories, including the RNA world hypothesis, metabolism-first scenarios, and protein-first models, treat the emergence of life as an improbable event requiring the fortuitous assembly of complex molecular structures under specific environmental conditions. These approaches implicitly assume that life represents a local thermodynamic improbability that arose through stochastic processes in a limited number of favorable locations. The electron transport partitioning framework fundamentally inverts this perspective by demonstrating that life's emergence was not improbable but thermodynamically inevitable given the ubiquity of electron-donating and electron-accepting chemical species in the early universe and the fundamental thermodynamic favorability of charge separation across boundaries.

The probability comparison established in Section~\ref{sec:orgels_paradox} demonstrates this inversion quantitatively through the ratio of membrane-first to information-first probabilities:
\begin{equation}
    \frac{P_{\text{membrane-first}}}{P_{\text{RNA-world}}} \approx \frac{10^{-6}}{10^{-150}} = 10^{144}
\end{equation}
This ratio approaches the estimated number of particles in the observable universe, indicating that membrane-first scenarios based on electron transport are not merely more probable than information-first scenarios but represent an entirely different class of thermodynamic processes. While information-first scenarios require the improbable assembly of specific sequences from vast combinatorial spaces, electron transport partitioning requires only the presence of redox-active species and geometric boundaries, both of which were abundant in prebiotic environments. The difference of 144 orders of magnitude places these scenarios in fundamentally distinct categories: information-first scenarios are thermodynamically negligible, while charge-partitioning scenarios are thermodynamically inevitable wherever appropriate chemical and physical conditions exist.

This shift from stochastic accident to thermodynamic inevitability has profound implications for our understanding of life's distribution in the universe. If life requires the improbable assembly of specific information-bearing sequences, then its occurrence may be rare even in environments that appear chemically favorable. However, if life emerges inevitably from electron transport partitioning wherever redox chemistry and geometric boundaries coexist, then life should be common throughout the universe, arising independently in any environment that supports charge separation. This prediction is testable through continued exploration of extreme environments on Earth and through the search for biosignatures on other planetary bodies.

\subsection{Resolution of the Homochirality Problem}

The universal homochirality of biological molecules—the exclusive use of L-amino acids in proteins, D-sugars in nucleic acids, and right-handed helical structures in DNA—has remained one of the most puzzling features of life, unexplained by information-first theories. Random chemical synthesis produces racemic mixtures containing equal proportions of left- and right-handed molecules, and there is no obvious chemical reason why life should exclusively use one enantiomer over the other. Information-first theories can describe how homochirality is maintained once established through template-directed replication, but they cannot explain how the initial chiral bias arose or why it is universal across all domains of life.

The electron transport partitioning framework provides a physical mechanism for the origin of homochirality that operates at the level of fundamental electromagnetic interactions. During electron transport through molecular structures, electron spin couples to the electromagnetic fields present in the transport environment, creating chiral selection pressures that favor one enantiomer over the other based on the relative efficiency of electron transport through molecules of different handedness. This chiral selection is not a property of individual molecules but emerges from the interaction between electron spin, molecular geometry, and electromagnetic field configuration. Once established in the first autocatalytic electron transport systems, this chiral preference propagates through geometric apertures at all subsequent organizational levels, from molecular recognition to macromolecular assembly to cellular structure.

This mechanism explains not only \emph{why} life is homochiral—because electron transport efficiency depends on molecular handedness—but also \emph{which} chirality was selected. The specific handedness observed in terrestrial biology (L-amino acids, D-sugars, right-handed DNA helices) was determined by the local electromagnetic field configuration during the establishment of the first autocatalytic electron transport systems on early Earth. This prediction suggests that if life arose independently in environments with different electromagnetic field configurations, it might exhibit opposite chirality, providing a potential biosignature for detecting independent origins of life.

\subsection{Interstellar Chemistry Explained}

The observation of complex organic molecules, including amino acids, nucleobases, and other biomolecular precursors, in cold and highly irradiated interstellar environments has presented a significant paradox for traditional kinetic theories of chemistry. In these environments, temperatures are typically 10-50 K, far below the activation energies required for conventional thermal chemistry, yet complex molecular synthesis clearly occurs. Meteoritic and cometary samples contain diverse organic compounds with structural complexity that should be impossible to achieve through thermal activation alone, and spectroscopic observations detect organic molecules in interstellar dust clouds where thermal reaction rates would be negligibly slow.

The geometric partitioning framework resolves this paradox by demonstrating that aperture-based molecular selection and assembly operate independently of temperature, depending only on molecular configuration and charge distribution rather than on thermal velocity distributions. Mineral grain surfaces in interstellar environments, particularly iron-sulfur minerals and phyllosilicate clays, function as semiconductor aperture arrays that facilitate electron transport and charge-driven molecular assembly. Photoexcitation from stellar radiation and ionization from cosmic rays create charge carriers in these semiconducting minerals, establishing electric fields that drive molecular synthesis through geometric partitioning rather than thermal activation. This mechanism explains how complex organic chemistry occurs in environments traditionally considered impossible for chemical synthesis and suggests that prebiotic chemistry is not limited to warm, liquid-water environments but can proceed wherever semiconducting surfaces and radiation sources coexist.

\subsection{Relationship to Existing Theories}

The electron transport partitioning framework does not entirely replace existing origin of life theories but rather provides their thermodynamic foundation, reinterpreting the roles of the components emphasized in each theory. The RNA world hypothesis, which proposes that self-replicating RNA molecules preceded both DNA and proteins, correctly identifies RNA as a crucial component in early life but misidentifies its primary function. In the electron transport partitioning framework, RNA's role is recast as a charge capacitor and aperture scaffold whose polyanionic backbone stabilizes cellular electric fields, rather than as a primordial information carrier. The catalytic activity of ribozymes, rather than representing primitive metabolism, reflects the ability of structured RNA to create geometric apertures that facilitate electron transport and charge-driven molecular selection.

Metabolism-first theories, which propose that autocatalytic chemical networks arose before genetic polymers, correctly emphasize the importance of energy flow but misidentify the source of this energy. In the electron transport partitioning framework, metabolism emerges as energy harvesting from electron transport gradients, not as the primordial system itself. The autocatalytic cycles proposed in metabolism-first scenarios, such as the reductive citric acid cycle, are reinterpreted as electron transport pathways that became elaborated and optimized through the evolution of protein catalysts, rather than as self-organizing chemical networks that preceded electron transport.

Lipid world scenarios, which suggest that self-assembling amphiphilic boundaries provided the first compartmentalization, correctly identify membranes as essential but misunderstand their primary function. In the electron transport partitioning framework, membranes are understood as electron transport scaffolds whose insulating properties enable charge separation, rather than as compartmentalization structures whose primary function is to separate interior from exterior. The negative charge of biological membranes, arising from phospholipid head groups, is not incidental but fundamental to their electron transport function, creating the electrostatic environment necessary for efficient charge partitioning.

Finally, iron-sulfur world theories, which propose that FeS clusters on mineral surfaces catalyzed early metabolism, are most consistent with the electron transport partitioning framework. FeS clusters are indeed early electron transport systems, and their ubiquity in modern metabolism reflects their role as primordial electron carriers. However, rather than functioning primarily as metabolic catalysts, FeS clusters in the electron transport framework are understood as the first autocatalytic electron transport systems, where electron movement through iron-sulfur centers created the electrochemical conditions necessary for further electron transport, initiating the self-referential dynamics that characterize life.

\section{Conclusion}
\label{sec:conclusion}

This work has established a comprehensive framework for understanding the origin of life through electron transport partitioning, resolving long-standing paradoxes and providing quantitative predictions that distinguish it from traditional information-first theories. We have demonstrated that Orgel's paradox—the circular dependency between genetic information, enzymatic catalysis, and metabolic energy production—is resolved by identifying electron transport partitioning as the more fundamental operation that precedes and enables all three components. Rather than requiring the simultaneous emergence of information, catalysis, and metabolism, life emerges from the simpler and thermodynamically favorable process of charge separation across geometric boundaries.

Our probability analysis has established that information-first scenarios are mathematically impossible when compared to membrane-first scenarios based on electron transport. The RNA world hypothesis, with an estimated probability of approximately $10^{-150}$ for the spontaneous assembly of a minimal self-replicating ribozyme, and DNA-first scenarios, with probabilities approaching $10^{-200}$ for the assembly of functional genetic systems, represent thermodynamic impossibilities that could not occur even given the entire age and volume of the observable universe. In contrast, membrane-first scenarios based on the self-assembly of amphiphilic molecules and the establishment of redox gradients have probabilities of approximately $10^{-6}$, representing events that would occur readily in prebiotic environments. The probability ratio exceeding $10^{144}$ places these scenarios in fundamentally different thermodynamic categories, establishing electron transport partitioning as the inevitable pathway for life's origin.

We have proven that autocatalytic electron transport systems—defined as systems where electron movement creates the electrochemical conditions necessary for further electron movement—represent the minimal self-referential structure capable of initiating biological complexity. Unlike information-based self-replication, which requires complex molecular machinery for template copying and error correction, autocatalytic electron transport requires only redox-active chemical species and geometric boundaries. This self-referential property allows electron transport systems to perpetuate and elaborate themselves without requiring external information templates or pre-existing catalytic machinery, providing a thermodynamically plausible pathway from non-living chemistry to living systems.

The framework of geometric partitioning has demonstrated that electron transport creates categorical apertures through charge field geometry, enabling molecular selection that operates independently of temperature or kinetic factors. This temperature-independent selection resolves the paradox of complex organic molecule formation in cold interstellar environments and explains how molecular specificity can arise without requiring the complex binding sites of evolved proteins. Geometric apertures created by charge fields select molecules based on their configuration and charge distribution, providing a physical mechanism for the molecular recognition that underlies all biological processes.

The universal homochirality of biological molecules—including L-amino acids in proteins, D-sugars in nucleic acids, and right-handed helical structures in DNA—has been established as direct evidence of partitioning primacy rather than as an unexplained quirk of biochemistry. Chiral selection emerges naturally from electron transport in electromagnetic fields, where electron spin coupling creates differential transport efficiency for molecules of different handedness. This chirality, once established in the first autocatalytic electron transport systems, propagates through geometric apertures at all organizational levels, from molecular recognition to macromolecular assembly to cellular structure, explaining why homochirality is universal across all domains of life.

We have demonstrated that biological membranes evolved primarily as electron transport scaffolding rather than as compartmentalization structures, fundamentally reinterpreting their role in the origin and operation of life. The lipid bilayer provides an insulating barrier that enables charge separation, while embedded protein complexes facilitate directed electron flow, creating the charge-separated state that powers all cellular work. The negative charge of biological membranes, arising from phospholipid head groups and often enhanced by anionic lipids, is not incidental but fundamental to their electron transport function, creating the electrostatic environment necessary for efficient charge partitioning.

The evolution of DNA and RNA has been reinterpreted as the evolution of charge capacitors that optimize electrostatic integration within cells. The human genome, with approximately $6 \times 10^9$ base pairs and corresponding phosphate charges, stores approximately $10^{-12}$ joules of electrostatic energy, representing a significant fraction of total cellular energy. The regular helical structure of nucleic acids, their polyanionic backbone, and their association with cationic proteins create charge distributions that stabilize the cellular electric field. Information storage, in this framework, emerges as an evolutionary bonus of charge dynamics—a secondary function that became possible once charge-stabilizing polymers existed—rather than as the primary selective pressure driving the evolution of genetic systems.

Finally, we have explained the formation of complex organic molecules in cold interstellar environments through the semiconductor behavior of mineral surfaces, which function as temperature-independent aperture arrays. Photoexcitation and cosmic ray bombardment create charge carriers in semiconducting minerals such as iron-sulfur compounds and phyllosilicate clays, establishing electric fields that drive molecular synthesis through geometric partitioning rather than thermal activation. This mechanism resolves the long-standing paradox of prebiotic chemistry in environments where thermal reaction rates would be negligibly slow and suggests that the chemical precursors of life are widespread throughout the universe.

These results collectively establish that life's origin was not a stochastic accident requiring improbable molecular assembly but rather a thermodynamic inevitability determined by the fundamental physics of electron transport and charge partitioning. The transition from non-living to living systems emerges as a continuous physical process driven by charge dynamics, with information storage, catalysis, and metabolism arising as elaborations of the fundamental charge-partitioning operation. This framework provides a unified foundation for understanding not only the origin of life on Earth but also the potential for life throughout the universe, wherever electron transport and geometric boundaries coexist.


\bibliographystyle{plainnat}
\bibliography{references}

\end{document}

