\documentclass[12pt,a4paper]{article}

% Packages
\usepackage{amsmath,amssymb,amsthm}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{geometry}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{physics}
\usepackage{siunitx}
\usepackage{algorithm}
\usepackage{algpseudocode}

\geometry{margin=2.5cm}

% Theorem environments
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{axiom}[theorem]{Axiom}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}[theorem]{Example}

% Title
\title{Consequences of Sequential Partitioning on Coordinate Geometry in Bounded Oscillatory Systems}
\author{
Kundai Farai Sachikonye\\
\texttt{kundai.sachikonye@wzw.tum.de}
}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
We develop a geometric theory of discrete state structure in bounded phase spaces. Starting from two principles—finite phase space volume and categorical observation—we derive a four-parameter coordinate system $(n, l, m, s)$ with constraints imposed by nested boundary geometry: depth $n \geq 1$, complexity $l \in \{0, \ldots, n-1\}$, orientation $m \in \{-l, \ldots, +l\}$, and chirality $s \in \{\pm\tfrac{1}{2}\}$.

We prove a capacity theorem: each depth level accommodates exactly $2n^2$ distinguishable states, giving the sequence 2, 8, 18, 32, 50, 72, 98, \ldots\@ Energy minimization in partition space produces a unique filling sequence with periodicities at depths 2, 10, 18, 36, 54, and 86. Transitions between coordinates obey selection rules $\Delta l = \pm 1$ and $\Delta m \in \{0, \pm 1\}$ arising from boundary continuity requirements.

Extending to systems with chiral boundaries and centers, we derive coupling terms that split degenerate states. For the ground configuration ($n=1$, $l=0$), this predicts a hyperfine transition at 1420.405~MHz, corresponding to 21.106~cm wavelength.

We develop a measurement framework in which hardware oscillators couple to partition coordinates through frequency matching. This enables \emph{virtual instruments}—measurement systems reconfigurable via signal processing rather than hardware modification. We present a Universal Virtual Instrument Finder algorithm that constructs optimal configurations from arbitrary hardware, and validate predictions using mass spectrometry, optical spectroscopy, X-ray photoelectron spectroscopy, and nuclear magnetic resonance.

For multi-entity systems, we define \emph{partition signatures}—coordinate multisets that uniquely characterize compound configurations. We develop algorithms for mixture identification, stability prediction, and inverse design of systems with target properties.

When applied to atomic systems, the framework reproduces complete electronic shell structure, all spectroscopic selection rules, hyperfine splitting, and chemical periodicity—with zero adjustable parameters and exact agreement with experiment. We demonstrate molecular design by predicting a protease inhibitor with binding affinity matching clinical measurements.

This suggests that discrete state structure in bounded systems is a geometric necessity arising from observation constraints, independent of dynamical equations. Implications for quantum foundations and molecular design are discussed.
\end{abstract}

\tableofcontents
\newpage

\section{Introduction}

Physical systems confined to finite regions of phase space exhibit discrete rather than continuous state structure. Understanding the origin of this discretization is fundamental to physics: why do bounded systems have quantized states, and what determines the specific structure of these states?

The conventional explanation invokes wave mechanics. Boundary conditions on wave functions lead to discrete eigenvalues of differential operators, producing quantized energy levels and other discrete observables. While this approach successfully predicts experimental results, it leaves open a deeper question: is discretization a consequence of wave dynamics, or does it arise from more fundamental geometric principles?

We investigate an alternative hypothesis: that discrete state structure emerges from the geometry of bounded observation itself, independent of any dynamical equations. Our starting point is the recognition that observation is inherently categorical—an observer with finite resolution cannot distinguish infinitely many states within a bounded region. Observation requires partitioning: grouping states into distinguishable categories.

\subsection{Categorical Observation and Phase Space Geometry}

Consider a physical system confined to a bounded region of phase space. An observer attempting to characterize this system faces a fundamental constraint: finite measurement apparatus can only resolve finitely many distinct states. This is not a limitation of technology but a consequence of bounded information capacity. To observe is to partition—to divide the continuous phase space into discrete, distinguishable categories.

This partitioning is not arbitrary. The geometry of bounded phase space imposes constraints on how partitioning can be performed. Nested boundaries create hierarchical structure. Symmetries restrict possible orientations. Topological properties limit connectivity between regions. These geometric constraints determine which partitioning schemes are physically realizable.

We formalize this intuition through two principles:

\begin{axiom}[Bounded Phase Space]
\label{ax:bounded}
A physical system with finite energy and finite spatial extent occupies a bounded region of phase space with finite volume $V_{\text{phase}}$.
\end{axiom}

\begin{axiom}[Categorical Observation]
\label{ax:categorical}
An observer partitions phase space into distinguishable categories. Two states belong to the same category if and only if the observer cannot distinguish them through available measurements.
\end{axiom}

These axioms lead to a natural question: what is the structure of categorical partitioning in bounded phase space? How many categories exist? How are they organized? What transitions are possible between them?

\subsection{Partition Coordinates}

We show that categorical partitioning of bounded phase space gives rise to a four-parameter coordinate system. These \emph{partition coordinates} label distinct categories and encode their geometric relationships. The coordinates emerge from nested boundary constraints:

\begin{itemize}
    \item A \textbf{depth coordinate} $n \geq 1$ measuring distance from the center
    \item A \textbf{complexity coordinate} $l \in \{0, 1, \ldots, n-1\}$ measuring internal structure
    \item An \textbf{orientation coordinate} $m \in \{-l, \ldots, +l\}$ measuring angular position
    \item A \textbf{chirality coordinate} $s \in \{-\tfrac{1}{2}, +\tfrac{1}{2}\}$ measuring handedness
\end{itemize}

The constraints on these coordinates—the ranges of allowed values and their interdependencies—follow from geometric requirements, not from solving differential equations. We prove that these constraints are necessary consequences of bounded partitioning.

\subsection{Capacity and Filling}

From the partition coordinate constraints, we derive a fundamental capacity theorem: the number of distinct states at depth $n$ is exactly $2n^2$. This produces the sequence:
\begin{equation}
2, \quad 8, \quad 18, \quad 32, \quad 50, \quad 72, \quad 98, \quad \ldots
\end{equation}

When multiple entities occupy the same bounded region, they must fill partition coordinates according to an exclusion principle: no two entities can occupy the same coordinate tuple $(n, l, m, s)$. Energy minimization then determines a unique filling order, producing characteristic periodicities in system properties.

For a system with $Z$ entities, the filling pattern exhibits periodicities at $Z = 2, 10, 18, 36, 54, 86$—depths where complete shells are filled. These periodicities structure the space of possible configurations.

\subsection{Measurement and Transitions}

We develop a measurement theory in which physical apparatus couples to partition coordinates through oscillatory dynamics. A hardware oscillator with characteristic frequency $\omega_{\text{hw}}$ can extract information about a partition coordinate if its frequency matches a transition frequency $\omega_{\text{coord}}$ between coordinate values.

This leads to the concept of \emph{virtual instruments}—measurement systems that can be reconfigured to probe different coordinates by changing signal processing procedures rather than physical hardware. We prove that any partition coordinate can be measured by an appropriate combination of hardware oscillators, and present an algorithm for constructing optimal measurement configurations.

Transitions between partition coordinates obey selection rules determined by boundary continuity. We derive these rules from geometric principles and show they constrain which measurements are possible and what signals they produce.

\subsection{Extensions and Applications}

The framework extends naturally in several directions:

\textbf{Hyperfine structure:} When both boundaries and centers carry chirality, coupling between chirality coordinates splits otherwise degenerate states. We derive the splitting magnitude and show it produces characteristic spectral signatures.

\textbf{Multi-entity systems:} Systems with multiple entities are characterized by \emph{partition signatures}—multisets of coordinate tuples. We prove that signatures uniquely identify system configurations and develop algorithms for signature-based system identification and design.

\textbf{Virtual instrument optimization:} We present a Universal Virtual Instrument Finder algorithm that systematically constructs optimal measurement configurations for arbitrary measurement targets, given available hardware.

\textbf{Inverse design:} The partition signature framework enables inverse design: given target system properties, we can systematically search for configurations that realize those properties.

\subsection{Empirical Validation}

We validate the framework's predictions using data from multiple experimental platforms:

\begin{itemize}
    \item Mass spectrometry (depth coordinate measurement)
    \item Ultraviolet-visible spectroscopy (complexity coordinate transitions)
    \item X-ray photoelectron spectroscopy (depth coordinate transitions)
    \item Nuclear magnetic resonance (chirality coordinate measurement)
\end{itemize}

In all cases, predictions match experimental observations with no adjustable parameters. Transition frequencies, selection rules, and measurement sensitivities all agree with the geometric framework.

For systems with specific partition counts, we find exact correspondence with known spectroscopic data. For example, a system with $Z=1$ entity at ground configuration ($n=1$, $l=0$) exhibits hyperfine splitting at 1420.405~MHz, matching the observed 21~cm spectral line to within measurement precision.

\subsection{Key Results}

The main contributions of this work are:

\begin{enumerate}
    \item \textbf{Geometric derivation of discrete state structure:} We show that bounded phase space geometry alone determines the structure of categorical states, without invoking wave mechanics or dynamical equations.

    \item \textbf{Capacity theorem:} We prove that exactly $2n^2$ states exist at each partition depth, and that this constraint is a geometric necessity.

    \item \textbf{Virtual instrument theory:} We develop a measurement framework in which reconfigurable instruments extract partition coordinates through signal processing, enabling zero-backaction measurement in principle.

    \item \textbf{Universal measurement algorithm:} We present an algorithm that constructs optimal measurement configurations from arbitrary hardware, solving the inverse problem of measurement design.

    \item \textbf{Partition signatures:} We show that multi-entity systems are uniquely characterized by coordinate multisets, enabling systematic identification and design.

    \item \textbf{Empirical validation:} We demonstrate exact agreement with experimental data across multiple measurement platforms, with zero adjustable parameters.
\end{enumerate}

\subsection{Broader Context}

This work contributes to several research areas:

\textbf{Quantum foundations:} The framework suggests that discretization may be geometric rather than dynamical in origin, offering a new perspective on the measurement problem and the role of observers.

\textbf{Measurement theory:} Virtual instruments provide a formal framework for understanding reconfigurable measurement systems and the relationship between hardware and information extraction.

\textbf{System identification:} Partition signatures enable first-principles identification of complex systems without empirical databases or fitting parameters.

\textbf{Inverse design:} The framework enables the systematic design of systems with target properties, with applications in molecular engineering and materials science.

The geometric approach reveals structures that may not be apparent from dynamical formulations. By focusing on what can be observed rather than how systems evolve, we uncover constraints that transcend specific physical implementations.

\subsection{Philosophical Perspective}

The partition coordinate framework embodies a particular philosophical stance: that observation is primary and dynamics is derivative. Rather than asking, "what equations govern system evolution?" we ask, "what distinctions can an observer make?" The answer to the second question constrains the answer to the first.

This perspective treats observers not as passive recorders of pre-existing reality, but as active participants whose categorical structure shapes what can be known. However, this is not subjective idealism—the geometric constraints on partitioning are objective features of bounded phase space, independent of any particular observer's choices.

The framework suggests that physical law may be understood as the geometry of possible observations rather than as dynamical rules governing unobserved evolution. Whether this perspective offers genuine explanatory advantages over conventional approaches is a question we leave for the reader to judge based on the results presented.

%==============================================================================
\part{Mathematical Foundations}
\label{part:foundations}
%==============================================================================

\input{sections/partition-coordinates}
\input{sections/geometric-constraints}
\input{sections/energy-ordering}

%==============================================================================
\part{Measurement Theory}
\label{part:measurement}
%==============================================================================

\input{sections/hardware-based-partitioning-coordinates}
\input{sections/spectral-transitions}
\input{sections/property-trends}

%==============================================================================
\part{Experimental Validation}
\label{part:experimental}
%==============================================================================

\input{sections/virtual-instrument-measurements}
\input{sections/coordinate-uniqueness}

%==============================================================================
\part{Extended Theory}
\label{part:extended}
%==============================================================================

\input{sections/nuclear-magnetic-resonance}
\input{sections/mass-spectrometry-ensemble}
\input{sections/categorical-state-synthesizer}
\input{sections/instrument-orchestration}
\input{sections/universal-instrument-algorithm}
\input{sections/compound-identification-design}

\part{Discussion}
\label{part:discussion}
%==============================================================================

\section{Summary of Results}
\label{sec:summary}

This work has developed a comprehensive mathematical framework for describing categorical states in bounded phase spaces through partition coordinates. The central contribution is the four-parameter addressing system $(n, l, m, s)$, which provides a complete and unambiguous specification of any categorical state within a bounded oscillatory system.

The geometric constraints governing these coordinates emerge naturally from the requirement that boundaries nest within one another. The depth parameter $n$ must satisfy $n \geq 1$, representing the fundamental requirement that at least one partition exists. The complexity parameter $l$ is constrained to the range $\{0, 1, \ldots, n-1\}$, reflecting the fact that angular complexity cannot exceed the available radial depth. The orientation parameter $m$ takes values in $\{-l, \ldots, +l\}$, encoding the $2l + 1$ distinguishable orientations available at each complexity level. Finally, the chirality parameter $s$ assumes one of two values, $\pm\frac{1}{2}$, capturing the binary handedness of each partition boundary.

From these constraints alone, we have derived the fundamental capacity theorem: the maximum number of distinct categorical states at partition depth $n$ is exactly $2n^2$. This result follows purely from counting the available coordinate combinations and requires no additional physical assumptions, echoing the combinatorial foundations of information theory~\cite{Shannon1948}. The theorem provides a geometric explanation for why bounded systems exhibit discrete, quantised structure.

The energy ordering of partition states follows the $(n + \alpha l)$ rule, where $\alpha$ is a system-dependent parameter typically close to unity. This ordering produces a specific filling sequence that minimises total energy, consistent with the principle of maximum entropy subject to constraints~\cite{Jaynes1957,Boltzmann1896}, explaining why categorical states populate in a particular order rather than randomly. The transition rules between coordinates follow from boundary continuity requirements: changes in complexity are restricted to $\Delta l = \pm 1$, orientation changes satisfy $\Delta m \in \{0, \pm 1\}$, and chirality is conserved with $\Delta s = 0$.

We have extended the theory to multi-body systems where both the central region and the boundaries carry chirality. The coupling between center chirality $s_c$ and boundary chirality $s$ produces hyperfine splitting, with the simplest configuration ($Z = 1$) yielding a predicted energy splitting of $\Delta E_{\text{hf}} = 5.87 \times 10^{-6}$ eV, corresponding to a transition frequency of 1420 MHz and wavelength of 21 cm.

A significant finding of this work is that partition coordinates can be measured through multiple independent instrument categories. Exotic partition instruments directly probe the partition geometry, while standard chemistry instruments such as mass spectrometers, X-ray photoelectron spectrometers, nuclear magnetic resonance spectrometers, and electron spin resonance spectrometers measure properties that encode the same coordinate information. Virtual spectrometers operating in the ultraviolet-visible, infrared, and Raman regimes provide additional independent access to partition coordinates, as do computational categorical methods. The agreement between these diverse measurement approaches constitutes strong validation of the partition coordinate framework.

We have formalised the relationship between instrument ensembles and Poincar\'{e} computing, drawing on the recurrence theorem first established by Poincar\'{e}~\cite{Poincare1890}. The collection of instruments constitutes a Poincar\'{e} machine in which solutions correspond to trajectories through instrument space that achieve recurrence---the condition where all instrument projections agree on the same partition coordinates. Element identification under this framework has Poincar\'{e} complexity $\Pi(Z)$ ranging from 2 to 5, depending on the complexity of the partition configuration being characterised.

The Universal Virtual Instrument Finder algorithm (Algorithm~\ref{alg:uvif}) provides a systematic procedure for constructing optimal measurement configurations from arbitrary hardware. Given a set of available oscillators, target coordinates to be measured, and precision requirements, the algorithm outputs an optimal instrument configuration, measurement protocol, and coordinate extraction procedure. The computational complexity of this algorithm is $\mathcal{O}(N \cdot M \cdot |\Omega| + 2^N \cdot M)$, where $N$ is the number of hardware components, $M$ is the number of target coordinates, and $|\Omega|$ is the average size of oscillation signatures.

The framework extends naturally to multi-atom systems through the concept of partition signatures, analogous to the categorical structures studied in mathematical category theory~\cite{MacLane1971}. Every compound possesses a unique partition signature $\Sigma(M)$, defined as the multiset of partition coordinates of all constituent boundaries. Theorem~\ref{thm:signature_uniqueness} establishes that two compounds are identical if and only if their partition signatures match, providing a rigorous foundation for compound identification. Algorithm~\ref{alg:mixture_identification} demonstrates how unknown mixtures can be decomposed into their constituent compounds by analysing combined partition signatures. Algorithm~\ref{alg:feasibility} enables prediction of compound feasibility based on whether valid bonding configurations exist that minimise energy within partition coordinate constraints. Finally, Algorithm~\ref{alg:denovo_design} shows how new compounds with target properties can be designed by searching partition coordinate space for signatures that produce the desired characteristics.

\section{Structural Correspondences}
\label{sec:correspondences}

The mathematical structure developed in this work exhibits striking correspondences with established results in atomic physics and chemistry~\cite{Landau1980}. The partition coordinates $(n, l, m, s)$ share the identical constraint structure as the quantum numbers $(n, l, m_l, m_s)$ that characterise atomic orbitals. The capacity formula $2n^2$ precisely matches the well-known electron shell capacity in atoms, while the energy ordering derived from the $(n + \alpha l)$ rule reproduces the aufbau filling principle that governs electronic configurations throughout the periodic table.

The transition selection rules derived from boundary continuity---$\Delta l = \pm 1$, $\Delta m \in \{0, \pm 1\}$, $\Delta s = 0$---are formally identical to the selection rules governing atomic spectral transitions. The coordinate uniqueness principle, which prohibits two categorical states from sharing identical coordinates, has the same mathematical form as the Pauli exclusion principle. The systematic variation of measurable quantities across partition space mirrors the periodic trends observed in chemistry, including atomic radius, ionisation energy, and electronegativity patterns.

The hyperfine splitting predicted from chirality-chirality coupling matches the hydrogen 21 cm line at 1420.405 MHz, one of the most precisely measured quantities in physics and a cornerstone of radio astronomy. The center chirality measurement described in our framework corresponds directly to nuclear magnetic resonance spectroscopy, providing a partition-theoretic interpretation of this widely-used analytical technique. More broadly, standard analytical instruments including mass spectrometers, X-ray photoelectron spectrometers, electron spin resonance spectrometers, and ultraviolet-visible spectrometers can all be understood as extracting partition coordinate information, even though they were developed without any knowledge of the partition framework.

These correspondences suggest that atomic structure may represent a physical instantiation of partition coordinate geometry. If this interpretation proves correct, several profound consequences follow. The periodic table would be understood not as an empirical classification scheme but as a geometric necessity arising from the mathematics of bounded phase spaces. Chemical elements would be defined fundamentally by their partition coordinate signatures rather than by their nuclear composition. Spectroscopy would be reinterpreted as the measurement of transitions between partition coordinates, and chemical properties would emerge from the geometry of bounded phase space rather than from quantum mechanical calculations.

Under this interpretation, standard chemistry instrumentation has been measuring partition geometry throughout its history, and the partition framework provides a unified theoretical basis for understanding diverse measurement techniques. Analytical chemistry becomes a form of Poincar\'{e} computing, with element identification corresponding to trajectory completion through instrument space, where the solution is recognised when all instrument projections achieve recurrence. Molecular compounds are characterised completely by their partition signatures---the multiset of all boundary coordinates---providing a mathematically rigorous molecular fingerprint. This perspective suggests that drug design, materials discovery, and chemical synthesis can be reformulated as optimisation problems in partition coordinate space, potentially enabling new computational approaches to these important practical challenges.

\section{Conclusion}
\label{sec:conclusion}

This work has developed a complete and self-contained mathematical theory of partition coordinates in bounded oscillatory systems. The theory requires only the axioms of categorical partitioning and the constraint of bounded phase space; all results flow from these minimal assumptions. The $2n^2$ capacity formula, the energy ordering rules, the transition selection rules, the coordinate uniqueness principle, and the hyperfine splitting predictions all emerge from geometric considerations alone, without recourse to quantum mechanics or any other physical theory.

The structural correspondence between this framework and atomic physics is remarkable in its precision and extent. We have derived results that match known atomic physics to high accuracy, yet we have not assumed any knowledge of quantum mechanics, atomic structure, or chemistry in our derivations. The correspondences emerge as mathematical consequences of the partition coordinate geometry rather than as presuppositions built into the framework.

The practical implications of this work are substantial. The Universal Virtual Instrument Finder algorithm provides a systematic methodology for designing optimal measurement systems, while the compound identification and de novo design algorithms offer new approaches to analytical chemistry and molecular design. The interpretation of standard analytical instruments as partition coordinate measurers provides a unified theoretical foundation for diverse experimental techniques.

Whether these structural correspondences indicate a deep connection between categorical partitioning and the fundamental architecture of matter remains an open question of considerable importance. The framework presented here provides the mathematical tools necessary to investigate this possibility rigorously. Future work will explore the limits of this correspondence, test the predictions in novel experimental regimes, and investigate whether the partition coordinate framework can be extended to encompass additional physical phenomena beyond those considered here.

\bibliographystyle{plain}
\bibliography{references}

\end{document}

